The ChEMBL database as Linked Open Data

### 2012-08-10 article.tex
The authors themselves spell it all upper case: SIDER
quick check for grammar on article

\begin{abstract}
\paragraph*{Background:}
Making data available as Linked Data using Resource Description Framework (RDF) promotes integration with other web resources.
RDF documents can natively link to related data, and others can link back using Universal Resource Identifiers (URIs).
RDF makes the data machine-readable and uses extensible vocabularies for additional information, making it easier
to scale up inference and data analysis.
\paragraph*{Results:}
This paper describes recent developments in an ongoing project converting data from the ChEMBL database into RDF triples.
Relative to earlier versions, this updated version of ChEMBL-RDF uses recently introduced ontologies, including CHEMINF and CiTO,
exposes more information from the database, and is now available as dereferencable, linked data.
To demonstrate these new features, we present novel use cases showing further integration with
other web resources, including Bio2RDF, Chem2Bio2RDF, and ChemSpider, and showing the use of standard
ontologies for querying.
\paragraph*{Conclusions:}
We have illustrated the advantage of using open standards and ontologies to link the ChEMBL database
to other databases. Using those links and the knowledge encoded in standards and ontologies, the ChEMBL-RDF
resource creates a foundation for integrated semantic web cheminformatics applications, 
such as the presented decision support.
\end{abstract}

@@

The current scientific data deluge, in which datasets grow faster in size and complexity than scientists can keep up with,
provides several new challenges for information systems. Scientists wish to 
discover new, unique, and significant patterns in datasets that explain biological 
phenomena not yet understood. This process requires integration of datasets that are both 
sparse, and growing independently of one another. The discovery of new patterns showing the 
causes and effects of various biological phenomena is often beyond the scope of single datasets.
For example, systems biologists integrate micro-array differential expression datasets to 
biological pathways, using various other datasets to provide evidence for the links~\cite{Staab2007}. 
Another prominent example is drug discovery, in which a new unique chemical entity is designed or discovered 
based on descriptions of its required biological properties. This process requires the effective 
linkage of many scientific datasets~\cite{Samwald2011,Williams2012}.

ChEMBL contains descriptions for biological activities involving over a million chemical 
entities, extracted primarily from the scientific literature, providing a unique resource for drug researchers~\cite{Gaulton2012,Warr2009}.
It is updated on a fairly frequent basis as the existing data is further curated and new data is added. 
The ChEMBL dataset is available for download and can be browsed using a web interface and web services.
The former requires scientists to import the data into a relational database, while the 
latter limits the machine access to the data. ChEMBL also provides web services for programmatic
access, but these are limited to specific types of query as encoded in the web service API.

Using the right to download, modify, and redistribute the ChEMBL data, two independent teams
have previously mapped the ChEMBL dataset to the Resource Description Framework (RDF), resulting
in Chem2Bio2RDF~\cite{Chen2010} and ChEMBL-RDF~\cite{Willighagen2011}. RDF is a framework where
knowledge is represented by small so-called triples, reflecting ``facts''. A fact can
be given on any topic, and RDF therefore talks about \textit{things} very generically~\cite{Miller:04:RP}. Things in RDF
are represented by Universal Resource Identifiers (URI) and when the more specialized
Universal Resource Locator (URL) is used, it creates \emph{linked data}~\cite{Samwald2011}. RDF is
developed as an open standard by the World Wide Web Consortium, and provides a knowledge
representation framework that applies to many, if not all, domains, making it a successful 
approach to data integration in the life sciences.

@@

It should also be noted that the relationship of these structural classes 
is that of having a \textit{role} of drug, rather than a subclass relationship.
Moreover, we triples only give this drug role to approved drugs, as defined by the
ChEMBL database. 
This is triplified using the OBO and ChEBI ontologies, where CHEBI\_23888 is the drug role:

@@

The Bio2RDF project provides both resolvable Linked Data URIs using a generic Linked Data
server~\cite{Ansell2011} to access SPARQL endpoints for a range of scientific databases~\cite{Belleau2008}.
A number of these databases are referenced in ChEMBL, including ChEBI, PubMed, and both the
UniProt protein and taxonomy databases~\cite{TheUniProtConsortium2010}. These links are
vital to provide context for use cases that require a correlation between chemical
structures and other scientific data.

@@

ChemSpider~\cite{Pence2010} is a freely accessible, online database provided by Royal Society of Chemistry (RSC).
It contains over 28 million unique
chemical compounds aggregated from over 400 data sources, as well as chemical data extracted
from RSC scientific articles and databases. Since its inception, efforts have been made to
utilize it as both a deposition platform for the community to contribute novel data, as
well as a platform for annotation and curation for existing data. Studies have shown that there are data
quality issues in many of the public compound databases~\cite{Williams2011}. ChemSpider has become a
valuable resource for curated data, especially chemical-compound name mappings. ChemSpider
is presently providing the chemical structure, substructure and similarity searching services 
underpinning the Open PHACTS semantic web project~\cite{Williams2012}. Specific chemical data 
sources containing data mappings between ChemSpider identifiers (CSIDs) and the original data 
source identifiers have been provided to the triple store, together with chemical identifiers 
including validated chemical names (systematic, generic and trivial), SMILES, and InChIs. 
The data mappings between the CSIDs and ChEMBL IDs are released to the 
community under the Creative Common Attribution-Share Alike license (CC-BY-SA 3.0). Attribution 
should be made to the original ChEMBL database, Open PHACTS, and ChemSpider.

@@

A second resource in the linked open data network into which ChEMBL-RDF has been integrated is Chem2Bio2RDF.
Chem2Bio2RDF is a single RDF repository covering over twenty public data resources pertaining to drugs, chemical compounds, carcinogens,
protein targets, genes, diseases, side effects, pathways and their relations~\cite{Chen2010}. The entities and their relations were
further annotated by Chem2Bio2OWL ontology, making it a rich semantic resource for integrative searches~\cite{Chen2012a} and
data mining~\cite{Chen2012b}. The ChEMBL-RDF set has been uploaded into the Chem2Bio2RDF triple store, enabling
queries linking ChEMBL with other entities in Chem2Bio2RDF. Since both ChEMBL-RDF and Chem2Bio2RDF use InChI keys to present chemical
compounds and adopt Bio2RDF protein identifiers to present targets, queries can be easily constructed to link the two. For instance,
to investigate the relations between drug side effects and their targets (usually off-target effects), a query was created to link the
targets in ChEMBL to side effects (e.g., heart disease) in Chem2Bio2RDF via side effect related drugs and their bioassay activities.
In this case, 36 drugs causing heart disease were linked to 87 unique protein targets ($IC_50$ $<$ 10um). The top two most common targets
of these drugs are HERG and Serotonin transporter. HERG is a well-known off-target accounting for cardiovascular problems, while
the activity of Serotonin transporter may also be associated with significant cardiovascular adverse effects~\cite{Levy2006}. 
The targets can be also linked to side effects via protein-protein interactions by searching across ChEMBL, HPRD and Sider in
Chem2Bio2RDF. SPARQLs are available at \url{http://chem2bio2rdf.wikispaces.com/Integrate+with+ChEMBL} and can be executed in the
SPARQL endpoint~\cite{Chen2010}.

>>>>

\begin{abstract}
\paragraph*{Background:}
Making data available as Linked Data using Resource Description Framework (RDF) promotes integration with other web resources.
RDF documents can natively link to related data, and others can link back using Universal Resource Identifiers (URIs).
RDF makes the data machine-readable and uses extensible vocabularies for additional information, making it easier
to scale up inference and data analysis.
\paragraph*{Results:}
This paper describes recent developments in an ongoing project converting data from the ChEMBL database into RDF triples.
Relative to earlier versions, this updated version of ChEMBL-RDF uses recently introduced ontologies, including CHEMINF and CiTO;
exposes more information from the database; and is now available as dereferencable, linked data.
To demonstrate these new features, we present novel use cases showing further integration with
other web resources, including Bio2RDF, Chem2Bio2RDF, and ChemSpider, and showing the use of standard
ontologies for querying.
\paragraph*{Conclusions:}
We have illustrated the advantage of using open standards and ontologies to link the ChEMBL database
to other databases. Using those links and the knowledge encoded in standards and ontologies, the ChEMBL-RDF
resource creates a foundation for integrated semantic web cheminformatics applications, 
such as the presented decision support.
\end{abstract}

@@

The current scientific data deluge, in which datasets grow faster in size and complexity than scientists can keep up with,
provides several new challenges for information systems. Scientists wish to 
discover new, unique, and significant patterns in datasets that explain biological 
phenomena not yet understood. The discovery of new patterns showing the 
causes and effects of various biological phenomena is often beyond the scope of single datasets.
For example, systems biologists integrate micro-array differential expression datasets to 
biological pathways, using various other datasets to provide evidence for the links~\cite{Staab2007}. 
Another prominent example is drug discovery, in which a new unique chemical entity is designed or discovered 
based on descriptions of its required biological properties. This process requires the effective 
linkage of many scientific datasets that are both sparse, and growing independently of one another~\cite{Samwald2011,Williams2012}.

ChEMBL contains descriptions for biological activities involving over a million chemical 
entities, extracted primarily from scientific literature. This provides a unique resource for drug researchers~\cite{Gaulton2012,Warr2009}.
It is updated on a fairly frequent basis as the existing data is further curated and new data is added. 
The ChEMBL dataset is available for download and can be browsed using a web interface and web services.
The former requires scientists to import the data into a relational database, while the 
latter limits the machine access to the data. ChEMBL also provides web services for programmatic
access, but these are limited to the specific types of queries which are encoded in the web service API.

Using the right to download, modify, and redistribute the ChEMBL data, two independent teams
have previously mapped the ChEMBL dataset to the Resource Description Framework (RDF), resulting
in Chem2Bio2RDF~\cite{Chen2010} and ChEMBL-RDF~\cite{Willighagen2011}. RDF is a framework where
knowledge is represented by small so-called triples, reflecting ``facts''. A fact can
be given on any topic, and RDF therefore talks about \textit{things} very generically~\cite{Miller:04:RP}. Things in RDF
are represented by Universal Resource Identifiers (URI). When the URI can be resolved, 
for example using the HTTP protocol, it creates \emph{linked data}~\cite{Samwald2011}. RDF is
developed as an open standard by the World Wide Web Consortium, and provides a knowledge
representation framework that applies to many, if not all, domains, making it a successful 
approach to data integration in the life sciences.

@@

It should also be noted that the relationship of these structural classes 
is that of having a \textit{role} of drug, rather than a subclass relationship.
Moreover, the triples only attach this drug role to approved drugs, as defined by the
ChEMBL database. 
This is triplified using the OBO and ChEBI ontologies in the following manner, where CHEBI\_23888 is the drug role:

@@

The Bio2RDF project provides both resolvable Linked Data URIs using a generic Linked Data
server~\cite{Ansell2011} and SPARQL endpoints containing RDF data from a range of scientific databases~\cite{Belleau2008}.
A number of these databases are referenced in ChEMBL, including ChEBI, PubMed, and both the
UniProt protein and taxonomy databases~\cite{TheUniProtConsortium2010}. These links are
vital to provide context for use cases that require a correlation between chemical
structures and other scientific data.

@@

ChemSpider~\cite{Pence2010} is a freely accessible, online database provided by Royal Society of Chemistry (RSC).
It contains over 28 million unique
chemical compounds aggregated from over 400 data sources, as well as chemical data extracted
from RSC scientific articles and databases. Since its inception, efforts have been made to
utilize it as both a deposition platform for the community to contribute novel data, as
well as a platform for annotation and curation for existing data. ChemSpider has become a
valuable resource for curated data, especially chemical-compound name mappings~\cite{Williams2011}. ChemSpider
is presently providing the chemical structure, substructure and similarity searching services 
underpinning the Open PHACTS semantic web project~\cite{Williams2012}. Specific chemical data 
sources containing data mappings between ChemSpider identifiers (CSIDs) and the original data 
source identifiers have been provided to the triple store, together with chemical identifiers 
including validated chemical names (systematic, generic and trivial), SMILES, and InChIs. 
The data mappings between the CSIDs and ChEMBL IDs are released to the 
community under the Creative Commons Attribution, Share Alike, license (CC-BY-SA 3.0). Attribution 
should be made to the original ChEMBL database, Open PHACTS, and ChemSpider.

@@

A second resource in the linked open data network into which ChEMBL-RDF has been integrated is Chem2Bio2RDF.
Chem2Bio2RDF is a single RDF repository covering over twenty public data resources pertaining to drugs, chemical compounds, carcinogens,
protein targets, genes, diseases, side effects, pathways and their relations~\cite{Chen2010}. The entities and their relations were
further annotated by Chem2Bio2OWL ontology, making it a rich semantic resource for integrative searches~\cite{Chen2012a} and
data mining~\cite{Chen2012b}. The ChEMBL-RDF set has been uploaded into the Chem2Bio2RDF triple store, enabling
queries linking ChEMBL with other entities in Chem2Bio2RDF. Since both ChEMBL-RDF and Chem2Bio2RDF use InChI keys to present chemical
compounds and adopt Bio2RDF protein identifiers to present targets, queries can be easily constructed to link the two. For instance,
to investigate the relations between drug side effects and their targets (usually off-target effects), a query was created to link the
targets in ChEMBL to side effects (e.g., heart disease) in Chem2Bio2RDF via side effect related drugs and their bioassay activities.
In this case, 36 drugs causing heart disease were linked to 87 unique protein targets ($IC_50$ $<$ 10um). The top two most common targets
of these drugs are HERG and Serotonin transporter. HERG is a well-known off-target accounting for cardiovascular problems, while
the activity of Serotonin transporter may also be associated with significant cardiovascular adverse effects~\cite{Levy2006}. 
The targets can be also linked to side effects via protein-protein interactions by searching across ChEMBL, HPRD and SIDER in
Chem2Bio2RDF. SPARQLs are available at \url{http://chem2bio2rdf.wikispaces.com/Integrate+with+ChEMBL} and can be executed in the
SPARQL endpoint~\cite{Chen2010}.
